---
title: Exploratory Data Analysis and Normalization
vignette: >
  % \VignetteIndexEntry{EDA}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    mathjax: null   
---

# Setup

```{r setup, message = FALSE, warning=FALSE}
library(SummarizedExperiment)
library(airway)
library(edgeR)
library(EDASeq)
library(ggplot2)
theme_set(theme_bw())
```

- TRADURRE IN INGLESE TUTTO

# The Airway data

```{r}
data(airway)
airway
colData(airway)
```

# Gene filtering

In this dataset, we have measurements for `r nrow(airway)` genes. Not all of them will be expressed in this particular system.

```{r}
table(rowSums(assay(airway))>0)
```

In general, it is advisable to remove those genes with a low expression level. Often, we may consider a filter that keeps those genes that are expressed in at least some reads in a certain number of samples.

The `filterByExpr` function of the `edgeR` package does that, considering by default 10 reads in at least 10 samples.

```{r}
filter <- filterByExpr(airway)
table(filter)
filtered <- airway[filter,]
filtered
```

# Exploratory Data Analysis

Exploratory Data Analysis (EDA) is useful to highlight those genes that have lower quality or that can have a high leverage in the downstream analyses.

## Total number of reads

First, we need to compute the total number of mapped reads per sample.

```{r}
airway$sum <- colSums(assay(airway))

df <- as.data.frame(colData(airway))

ggplot(df, aes(x=SampleName, y=sum, fill=cell)) +
    geom_bar(stat="identity")

ggplot(df, aes(x=SampleName, y=sum, fill=dex)) +
    geom_bar(stat="identity")
```

While there is clearly a difference in the total number of reads, it does not appear to be related to either cell type or treatment.

## Count distribution

Another immediately available plot is the boxplot of the count distribution.
Given the skewness of count distributions, we typically perform a log transformation.

```{r}
pal <- palette()[-1]
boxplot(log1p(assay(filtered)), col=pal[filtered$cell], las=2)
```

Se vogliamo evitare di calcolare i logaritmi per ogni grafico, possiamo salvare la matrice dei logaritmi in un nuovo slot dell'oggetto `filtered`.

```{r}
assay(filtered, "logcounts") <- log1p(assay(filtered))
filtered
```

## Relative Log Expression (RLE)

Come detto a lezione, la distribuzione RLE è spesso più utile di quella dei conteggi, specialmente per confrontare le normalizzazioni.

Il pacchetto `EDASeq` contiene la funzione `plotRLE` che può essere usata in questo caso.

```{r}
plotRLE(assay(filtered), col=pal[filtered$cell], outline=FALSE, las=2)
```

## Analisi delle componenti principali

Osservare come le prime due o tre componenti principali segregano i campioni può aiutare a capire quali sono le variabili che più influenzano i valori di espressione. Idealmente, vorremmo che i campioni si raggruppassero per classe biologica.

Il pacchetto `EDASeq` ha una funzione `plotPCA` che può essere usata in questo caso. In alternativa la PCA può essere calcolata manualmente con la solita funzione `prcomp()`.

```{r}
tmp <- assay(filtered)
colnames(tmp) <- filtered$cell
plotPCA(tmp, col=pal[filtered$dex])
```

# Normalization

```{r}
assay(filtered, "uq") <- betweenLaneNormalization(assay(filtered), which="upper")
```

```{r}
boxplot(log1p(assay(filtered, "uq")), col=pal[filtered$cell], las=2)
```

```{r}
plotRLE(assay(filtered, "uq"), col=pal[filtered$cell], las=2, outline=FALSE)
```

```{r}
tmp <- assay(filtered, "uq")
colnames(tmp) <- filtered$cell
plotPCA(tmp, col=pal[filtered$dex])
```

